Forcing for Isabelle

### 2018-08-24 blueprint.tex
typos y adecuación con la notación ya implementada

\begin{definition}
  \begin{enumerate}
  \item The relation \verb|eee|$(M)$ is the restriction to $M$ of the
    relational composition 
    ${\in}\circ{\in}\circ{\in}$. 
  \item For $\tau\in M$, the \emph{$G$-value} or \emph{interpretation
    under $G$} of $\tau$ is
    \[
    \val(\tau,\PP,G)\defi \{\val(\sigma,\PP,G) : \exists p\in \PP
    (\lb\sigma,p\rb\in\tau \y p\in G)\}.
    \]
  \end{enumerate}
\end{definition}

The definition of $\val$ proceeds by well-founded recursion on
\verb|eee|$(M)$. Determining  $\val(\tau,\PP,G)$ uses the values
$\val(\sigma,\PP,G)$ for smaller $\sigma$:
\[
\sigma\in\{\sigma\}\in\lb\sigma,p\rb\in\tau.
\]

@@

\begin{definition}
  The \emph{extension of $M$ by $G$} is
  \[
  \verb|gen_ext|(M,\PP,G) = M[G] \defi \{\val(\tau,\PP,G): \tau\in M\}.
  \]
\end{definition}
The standard definition of $M[G]$ adds the clause that $\tau$ should be
a $\PP$-name, but this is not necessary. When $G$ is $M$-generic for
$\PP$, we say that $M[G]$ is a \emph{generic extension}.

@@

\section{The fundamental theorems}
These are stated as two lemmas: \emph{Truth} and
\emph{Definability}. The second one ``simply'' says that the forcing
relation is definable. The previous recursive construction amounts to
that. The Truth Lemma states that the forcing relation indeed relates
satisfaction in $M[G]$ to that in $M$. 
\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
  language of set 
  theory with all free variables displayed. Then the
  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
  \begin{enumerate}
  \item     $M[G]\models\phi(\val(\tau_1,G),\dots,\val(\tau_n,G))$ for
    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
  \item $M\models\forceisa(\phi(\tau_1,\dots,\tau_n),\PP,\leq,\1,p)$.
  \end{enumerate}
\end{lemma}
A bit more formally, one should write the parameters to the left of
$\models$:
\begin{gather*}
  \forall G : p\in G \limp M[G],[\val(\tau_1,G),\dots,\val(\tau_n,G)]
  \models\phi(x_1,\dots,x_n) \\
  \text{if and only if}\\
  M,[\tau_1,\dots,\tau_n,\PP,\leq,\1,p] 
  \models\forceisa(\phi(x_1,\dots,x_n),x_{n+1},x_{n+2},x_{n+3},x_{n+4})
\end{gather*}
The proof of the Truth Lemma is by recursion in elements of
$\formula$. The following auxiliary results (adapted from
\cite[IV.2.43]{kunen2011set}) are also proved by recursion in
$\formula$.
\begin{lemma}\label{lem:strengthen} 
  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}.
  $M\models\forceisa(\phi,\PP,\leq,\1,p)$ and $p_1\leq p$
  implies $M\models\forceisa(\phi,\PP,\leq,\1,p_1)$.
\end{lemma}
\begin{lemma}\label{lem:density}
  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}. $M\models\forceisa(\phi,\dots)$ if and only if
    $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
    dense below $p$.''
\end{lemma}
\begin{lemma}\label{lem:forcing-negation}
  Under the same hypothesis of Lemma~\ref{lem:truth-lemma}, we have
  $M\models\forceisa(\phi,\dots)$ if and only if 
  $M\models\neg\exists
  q\leq p \,(\neg \forceisa(\phi,\PP,\leq,\1,q))$.
\end{lemma}

@@

We restate Lemma~\ref{lem:truth-lemma} in an alternative way, for
atomic formulas.
\begin{lemma}
  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
  in $M$, $p\in\PP$, $G\sbq\PP$ $M$-generic for $\PP$ and $\phi(x_1,x_2)$ an
  atomic formula. Then the following hold for all
  $\tau_1,\tau_2\in M$:
  \begin{enumerate}
  \item If $p\in G$ and
    \begin{equation} \label{eq:M-sats-forces-atom}
      M,[\tau_1,\tau_2,\PP,\leq,\1,p] 
      \models\forceisa(\phi(x_1,x_2),x_{3},x_{4},x_{5},x_{6})
    \end{equation}
    then
    \begin{equation} \label{eq:MG-sats-atom}
      M[G],[\val(\tau_1,G),\val(\tau_2,G)]
      \models\phi(x_1,x_2).
    \end{equation}
  \item If (\ref{eq:MG-sats-atom}) holds 
    then there is some $p\in G$ such that
    (\ref{eq:M-sats-forces-atom}) holds for that $p$.
  \end{enumerate}
\end{lemma}
This proof consists of two cases for $\phi$ (namely, $x_1\in x_2$ and
$x_1=x_2$) and it is done by well-founded recursion over $R$ (defined
in Section~\ref{sec:forcing-atomic-formulas}).

@@

\begin{lemma}
  Let $G$ be $M$-generic for $\PP$. For every $\phi\in\formula$ with
  three free variables, the $\phi$-instance of Separation holds in
  $M[G]$.
\end{lemma}
\begin{proof}
  Let $G$ as in the hypothesis, $\phi(x,z,v)$ be in \formula{} and let
  $c,w\in M[G]$. Following Kunen \cite[Lemma IV.2.6]{kunen2011set},
  we have to show that 
  \[
  S\defi\{x\in c : M[G]\models \phi(x,c,w)\} \in M[G].
  \]
  From this, using absoluteness, we will be able to derive the
  $\phi$-instance of Separation. Let $\pi,\sig\in M$ such that 
  $\val(\pi)=c$ and $\val(\sig)=w$.
  We will show that the name
  \[
  \tau \defi \{\lb\th,p\rb: \th\in\dom(\pi) \y p\in \PP\y
  M\models p\forces(\th\in\pi\y\phi(\th,\pi,\sig)\}
  \]
  belongs to $M$; for this we will use that an appropriate instance of
  Separation holds in $M$. Later we will see that $\tau$ names $S$.

@@

  Considering that only a fraction of elements
  of the environment  are necessary to interpret
  $\forceisa(\dB{7},\dB{8},\dB{9},\dB{1},\dB{0}\in\pi\y\phi(\dB{0},\dB{5},\dB{6}))$ and by permuting, we can
  replace
  \[
  M, [\th,p,x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1]\models
  \forceisa(\dB{7},\dB{8},\dB{9},\dB{1},\dB{0}\in \dB{5} \y\phi(\dB{0},\dB{5},\dB{6}))
  \]
  by
  \[
  M,[\PP,\leq,\1,p,\th,\pi,\sig]\models
  \forceisa(\dB{0},\dB{1},\dB{2},\dB{3},\dB{4}\in \dB{5} \y\phi(\dB{4},\dB{5},\dB{6}))
  \]
  By the Truth Lemma, this is equivalent to
  \[
  \forall F\ \PP\text{-generic}, p\in F \limp
  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
  \dB{0}\in\dB{1}  \y\phi(\dB{0},\dB{1},\dB{2}).
  \]

  Now we calculate $\val(\tau,F)$.
  \begin{multline*}
    \val(\tau,G) = \\
    \val\bigl(\{x\in\dom(\pi)\times\PP : \exists \th,p : x =\lb\th,p\rb
    \y  \\
    \forall F\ \PP\text{-generic}, p\in F \limp
    M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
    \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}\bigr).
  \end{multline*}
  Hence
  \begin{multline}\label{eq:val_tau}
    \val(\tau,G) = 
    \{\val(\th,G): \th\in\dom(\pi) \y  p \in G 
    \y \forall F\ \PP\text{-generic}, \\
   p\in F \longrightarrow
  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
  \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
  \end{multline}
  This is included in
  \begin{multline*}
    \{\val(\th,G): \th\in\dom(\pi) \y  p \in G 
    \y  \\
   p\in G \longrightarrow
   M[G],[\val(\th,G),\val(\pi,G),\val(\sig,G)]\models
   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\},
  \end{multline*}
  and in turn,  replacing $\val(\pi,F),\val(\sig,F)$ by their values,
  this is included  in 
  \[ 
    \{\val(\th,G): \th\in\dom(\pi)
    \y 
   M[G],[\val(\th,G),c,w]\models
   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
  \]
  By absoluteness, we have that $\val(\tau,G)$ is
  \[
    \{\val(\th,G): \th\in\dom(\pi)
    \y \val(\th,G)\in c \y 
   M[G],[\val(\th,G),c,w]\models
   \phi(\dB{0},\dB{1},\dB{2})\}.
  \]
  which equals 
  \[
    \{x: x\in c \y 
    M[G],[x,c,w]\models
   \phi(\dB{0},\dB{1},\dB{2})\}.
  \]
  since for every $x\in c=\val(\pi,G)$ there must
  exist some $\th\in\dom(\pi)$ with $\val(\th,G)=x$. Therefore
  $\val(\tau,G)\subseteq S$.

  Now let $x\in S$. Hence there exists $\lb\th,q\rb\in\pi$ such that
  $q\in G$ and $x=\val(\th,G)$. By the second version of the Truth
  Lemma, since 
  \[
  M[G],[\val(\th,G),\val(\pi,G),\val(\sig,G)]\models
  \dB{0}\in\dB{1}\y\phi(\dB{0},\dB{1},\dB{2}),
  \]
  there must exists $r\in G$ such that 
  \[
  M,[\PP,\leq,\1,r,\th,\pi,\sig]\models
  \forceisa(\dB{0},\dB{1},\dB{2},\dB{3},
  (\dB{4}\in\dB{5}\y\phi(\dB{4},\dB{5},\dB{6}))).
  \]
  Since $G$ is a filter, there is $p\in G$ such that $p\leq q, r$.
  By Lemma~\ref{lem:strengthen}, we have
  \[
  M,[\PP,\leq,\1,p,\th,\pi,\sig]\models
  \forceisa(\dB{0},\dB{1},\dB{2},\dB{3},
  (\dB{4}\in\dB{5}\y \phi(\dB{4},\dB{5},\dB{6}))),
  \]
  which by the Truth Lemma gives us
  \[
  \forall F\ \PP\text{-generic}, \\
   p\in F \longrightarrow
  M[F],[\val(\th,F),\val(\pi,F),\val(\sig,F)]\models
  \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2}).
  \]
  Using now (\ref{eq:val_tau}), this shows that $S\sbq \val(\tau,G)$,
  since $\th\in\dom(\pi)$. 
\end{proof}

>>>>

\begin{definition}
  \begin{enumerate}
  \item The relation \verb|eee|$(M)$ is the restriction to $M$ of the
    relational composition 
    ${\in}\circ{\in}\circ{\in}$. 
  \item For $\tau\in M$, the \emph{$G$-value} or \emph{interpretation
    under $G$} of $\tau$ is
    \[
    \val(\PP,G,\tau)\defi \{\val(\PP,G,\sigma) : \exists p\in \PP
    (\lb\sigma,p\rb\in\tau \y p\in G)\}.
    \]
  \end{enumerate}
\end{definition}

The definition of $\val$ proceeds by well-founded recursion on
\verb|eee|$(M)$. Determining  $\val(\PP,G,\tau)$ uses the values
$\val(\PP,G,\sigma)$ for smaller $\sigma$:
\[
\sigma\in\{\sigma\}\in\lb\sigma,p\rb\in\tau.
\]

@@

\begin{definition}
  The \emph{extension of $M$ by $G$} is
  \[
  \verb|gen_ext|(M,\PP,G) = M[G] \defi \{\val(\PP,G,\tau): \tau\in M\}.
  \]
\end{definition}
The standard definition of $M[G]$ adds the clause that $\tau$ should be
a $\PP$-name, but this is not necessary. When $G$ is $M$-generic for
$\PP$, we say that $M[G]$ is a \emph{generic extension}.

@@

\section{The fundamental theorems}
These are stated as two lemmas: \emph{Truth} and
\emph{Definability}. The second one ``simply'' says that the forcing
relation is definable. The previous recursive construction amounts to
that. The Truth Lemma states that the forcing relation indeed relates
satisfaction in $M[G]$ to that in $M$. 
\begin{lemma}[Truth Lemma]\label{lem:truth-lemma}
  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
  in $M$, $p\in\PP$, and $\phi(x_1,\dots,x_n)$ a formula in the
  language of set 
  theory with all free variables displayed. Then the
  following are equivalent, for all $\tau_1,\dots,\tau_n\in M$: 
  \begin{enumerate}
  \item     $M[G]\models\phi(\val(G,\tau_1),\dots,\val(G,\tau_n))$ for
    all filters $G$, $M$-generic for $\PP$, such that $p\in G$.
  \item $M\models\forceisa(\phi(\tau_1,\dots,\tau_n),\PP,\leq,\1,p)$.
  \end{enumerate}
\end{lemma}
A bit more formally, one should write the parameters to the left of
$\models$:
\begin{gather*}
  \forall G : p\in G \limp M[G],[\val(G,\tau_1),\dots,\val(G,\tau_n)]
  \models\phi(x_1,\dots,x_n) \\
  \text{if and only if}\\
  M,[\tau_1,\dots,\tau_n,\PP,\leq,\1,p] 
  \models\forceisa(\phi(x_1,\dots,x_n),x_{n+1},x_{n+2},x_{n+3},x_{n+4})
\end{gather*}
The proof of the Truth Lemma is by recursion in elements of
$\formula$. The following auxiliary results (adapted from
\cite[IV.2.43]{kunen2011set}) are also proved by recursion in
$\formula$.
\begin{lemma}\label{lem:strengthen} 
  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}.
  $M\models\forceisa(\phi,\PP,\leq,\1,p)$ and $p_1\leq p$
  implies $M\models\forceisa(\phi,\PP,\leq,\1,p_1)$.
\end{lemma}
\begin{lemma}\label{lem:density}
  Assume the same hypothesis of Lemma~\ref{lem:truth-lemma}. $M\models\forceisa(\phi,\dots)$ if and only if
    $M\models$``$\{p_1\leq p : \forceisa(\phi,\PP,\leq,\1,p_1)\}$ is
    dense below $p$.''
\end{lemma}
\begin{lemma}\label{lem:forcing-negation}
  Under the same hypothesis of Lemma~\ref{lem:truth-lemma}, we have
  $M\models\forceisa(\phi,\dots)$ if and only if 
  $M\models\neg\exists
  q\leq p \,(\neg \forceisa(\phi,\PP,\leq,\1,q))$.
\end{lemma}

@@

We restate Lemma~\ref{lem:truth-lemma} in an alternative way, for
atomic formulas.
\begin{lemma}
  Let $M$ be a ctm of $\ZFC$, $\lb\PP,\leq,\1\rb$ a forcing notion
  in $M$, $p\in\PP$, $G\sbq\PP$ $M$-generic for $\PP$ and $\phi(x_1,x_2)$ an
  atomic formula. Then the following hold for all
  $\tau_1,\tau_2\in M$:
  \begin{enumerate}
  \item If $p\in G$ and
    \begin{equation} \label{eq:M-sats-forces-atom}
      M,[\tau_1,\tau_2,\PP,\leq,\1,p] 
      \models\forceisa(\phi(x_1,x_2),x_{3},x_{4},x_{5},x_{6})
    \end{equation}
    then
    \begin{equation} \label{eq:MG-sats-atom}
      M[G],[\val(G,\tau_1),\val(G,\tau_2)]
      \models\phi(x_1,x_2).
    \end{equation}
  \item If (\ref{eq:MG-sats-atom}) holds 
    then there is some $p\in G$ such that
    (\ref{eq:M-sats-forces-atom}) holds for that $p$.
  \end{enumerate}
\end{lemma}
This proof consists of two cases for $\phi$ (namely, $x_1\in x_2$ and
$x_1=x_2$) and it is done by well-founded recursion over $R$ (defined
in Section~\ref{sec:forcing-atomic-formulas}).

@@

\begin{lemma}
  Let $G$ be $M$-generic for $\PP$. For every $\phi\in\formula$ with
  three free variables, the $\phi$-instance of Separation holds in
  $M[G]$.
\end{lemma}
\begin{proof}
  Let $G$ as in the hypothesis, $\phi(x,z,v)$ be in \formula{} and let
  $c,w\in M[G]$. Following Kunen \cite[Lemma IV.2.6]{kunen2011set},
  we have to show that 
  \[
  S\defi\{x\in c : M[G]\models \phi(x,c,w)\} \in M[G].
  \]
  From this, using absoluteness, we will be able to derive the
  $\phi$-instance of Separation. Let $\pi,\sig\in M$ such that 
  $\val(G,\pi)=c$ and $\val(G,\sig)=w$.
  We will show that the name
  \[
  \tau \defi \{\lb\th,p\rb: \th\in\dom(\pi) \y p\in \PP\y
  M\models p\forces(\th\in\pi\y\phi(\th,\pi,\sig)\}
  \]
  belongs to $M$; for this we will use that an appropriate instance of
  Separation holds in $M$. Later we will see that $\tau$ names $S$.

@@

  Considering that only a fraction of elements
  of the environment  are necessary to interpret
  $\forceisa(\dB{7},\dB{8},\dB{9},\dB{1},\dB{0}\in\pi\y\phi(\dB{0},\dB{5},\dB{6}))$ and by permuting, we can
  replace
  \[
  M, [\th,p,x,\tau,\dom(\pi)\times\PP,\pi,\sig,\PP,\leq,\1]\models
  \forceisa(\dB{7},\dB{8},\dB{9},\dB{1},\dB{0}\in \dB{5} \y\phi(\dB{0},\dB{5},\dB{6}))
  \]
  by
  \[
  M,[\PP,\leq,\1,p,\th,\pi,\sig]\models
  \forceisa(\dB{0},\dB{1},\dB{2},\dB{3},\dB{4}\in \dB{5} \y\phi(\dB{4},\dB{5},\dB{6}))
  \]
  By the Definition of Forcing, this is equivalent to
  \[
  \forall F\ \PP\text{-generic}, p\in F \limp
  M[F],[\val(F,\th),\val(F,\pi),\val(F,\sig)]\models
  \dB{0}\in\dB{1}  \y\phi(\dB{0},\dB{1},\dB{2}).
  \]

  Now we calculate $\val(G,\tau)$.
  \begin{multline*}
    \val(G,\tau) = \\
    \val\bigl(\{x\in\dom(\pi)\times\PP : \exists \th,p : x =\lb\th,p\rb
    \y \forall F\ \PP\text{-generic}, p\in F \limp \\    
    M[F],[\val(F,\th),\val(F,\pi),\val(F,\sig)]\models
    \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}\bigr).
  \end{multline*}
  Hence
  \begin{multline}\label{eq:val_tau}
    \val(G,\tau) = 
    \{\val(G,\th): \th\in\dom(\pi) \y  p \in G 
    \y \forall F\ \PP\text{-generic}, \\
   p\in F \longrightarrow
  M[F],[\val(F,\th),\val(F,\pi),\val(F,\sig)]\models
  \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
  \end{multline}
  This is included in
  \begin{multline*}
    \{\val(G,\th): \th\in\dom(\pi) \y  p \in G 
    \y  \\
   p\in G \longrightarrow
   M[G],[\val(G,\th),\val(G,\pi),\val(G,\sig)]\models
   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\},
  \end{multline*}
  and in turn,  replacing $\val(G,\pi),\val(G,\sig)$ by their values,
  this is included  in 
  \begin{equation}\label{eq:2}
    \{\val(G,\th): \th\in\dom(\pi)
    \y 
   M[G],[\val(G,\th),c,w]\models
   \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2})\}.
  \end{equation}
  By absoluteness, we have that (\ref{eq:2}) is the same as
  \[
    \{\val(G,\th): \th\in\dom(\pi)
    \y \val(G,\th)\in c \y 
   M[G],[\val(G,\th),c,w]\models
   \phi(\dB{0},\dB{1},\dB{2})\}.
  \]
  which equals 
  \[
    \{x: x\in c \y 
    M[G],[x,c,w]\models
   \phi(\dB{0},\dB{1},\dB{2})\}.
  \]
  since for every $x\in c=\val(G,\pi)$ there must
  exist some $\th\in\dom(\pi)$ with $\val(G,\th)=x$. Therefore
  $\val(G,\tau)\subseteq S$.

  Now let $x\in S$. Hence there exists $\lb\th,q\rb\in\pi$ such that
  $q\in G$ and $x=\val(G,\th)$. By the  Truth
  Lemma, since 
  \[
  M[G],[\val(G,\th),\val(G,\pi),\val(G,\sig)]\models
  \dB{0}\in\dB{1}\y\phi(\dB{0},\dB{1},\dB{2}),
  \]
  there must exist $r\in G$ such that 
  \[
  M,[\PP,\leq,\1,r,\th,\pi,\sig]\models
  \forceisa(\dB{0},\dB{1},\dB{2},\dB{3},
  (\dB{4}\in\dB{5}\y\phi(\dB{4},\dB{5},\dB{6}))).
  \]
  Since $G$ is a filter, there is $p\in G$ such that $p\leq q, r$.
  By Lemma~\ref{lem:strengthen}, we have
  \[
  M,[\PP,\leq,\1,p,\th,\pi,\sig]\models
  \forceisa(\dB{0},\dB{1},\dB{2},\dB{3},
  (\dB{4}\in\dB{5}\y \phi(\dB{4},\dB{5},\dB{6}))),
  \]
  which by the Truth Lemma gives us
  \[
  \forall F\ \PP\text{-generic}, \\
   p\in F \longrightarrow
  M[F],[\val(F,\th),\val(F,\pi),\val(F,\sig)]\models
  \dB{0}\in \dB{1} \y\phi(\dB{0},\dB{1},\dB{2}).
  \]
  Using now (\ref{eq:val_tau}), this shows that $S\sbq \val(G,\tau)$,
  since $\th\in\dom(\pi)$. 
\end{proof}
